### 2012-09-21 hdf5/discs_2012/abstract.tex
almost complete. yet to change figures

HDF5 is a data model, library and file format for storing and managing data. It is designed for flexible and efficient I/O for high volume and complex data. Natively, it uses a single-file format where multiple HDF5 objects are stored in a single file. In a parallel HDF5 application, multiple processes access a single file, thereby resulting in a performance bottleneck in I/O. Additionally, a single-file format does not allow semantic post processing on individual objects outside the scope of the HDF5 application. We have written a new plugin for HDF5 using its Virtual Object Layer (VOL) that provides serves two purposes: 1) it uses PLFS to convert the single-file layout into a data layout that is optimized for the underlying file system, and 2) it stores data in a unique way that enables semantic post-processing on the data, independently of HDF5. In this paper, we measure the performance improvements achieved using the plugin and discuss future work leveraging the new semantic post-processing functionality enabled.  We further discuss the applicability of this approach for exascale burst buffer storage systems.

>>>>

HDF5 is a data model, library and file format for storing and managing data. It is designed for flexible and efficient I/O for high volume and complex data. Natively, it uses a single-file format where multiple HDF5 objects are stored in a single file. In a parallel HDF5 application, multiple processes access a single file, thereby resulting in a performance bottleneck in I/O. Additionally, a single-file format does not allow semantic post processing on individual objects outside the scope of the HDF5 application. We have developed a new plugin for HDF5 using its Virtual Object Layer that provides serves two purposes: 1) it uses PLFS to convert the single-file layout into a data layout that is optimized for the underlying file system, and 2) it stores data in a unique way that enables semantic post-processing on the data. We measure the performance of the plugin and discuss work leveraging the new semantic post-processing functionality enabled.  We further discuss the applicability of this approach for exascale burst buffer storage systems.

### 2012-09-21 hdf5/discs_2012/background.tex
almost complete. yet to change figures

\subsection{HDF5}
HDF5 is a data model , library and file format for storing and managing data. It defines an information set, which is a container of array variables, groups, and types. They are known as datasets, groups and datatype objects in HDF5 respectively. The data model defines mechanisms for creating associations between various information items. The main components of HDF5 are briefly described below.

File:
In the HDF5 data model the container of an HDF5 infoset
is represented by an HDF5 file. It is a collection of HDF5 objects and represents the relationship between them.
Every file begins with a root group "/", which serves as the "starting-point" in the object hierarchy. An empty file has atleast the root group in it.

Datasets:
HDF5 datasets are objects that represent actual data or content. Datasets are arrays which can have multiple dimensions. A dataset is characterized by a dataspace and a datatype. An HDF5 dataspace captures the rank (number of dimensions) , and the current and maximum extent in the respective dimensions. An HDF5 datatype describes the type of its data elements. 
HDF5 supports ten classes of datatypes including scalar types like integer, floating-point, string, etc. and compound datatypes.

Groups:
HDF5 groups are akin to directories in a file system (reference here). A group is an explicit association between HDF5 objects. A group could contain multiple other groups, datasets or datatypes /textit{within} it. 
Traversing a root group generates the /information set graph/.

Attributes:
Aattributes are used for annotating datasets, groups, and datatype objects. They are datasets themselves, and are /textit{attached} to the existing objects they annotate.

The above definitions are explained in the example shown in figure ~\ref{hdf5_example}. 
The file "sample.h5" contains a root group which itself contains a group G1 and two datasets, D1 and D2. Group G1 contains a dataset D3. Attribute A1 is linked to dataset D1. The datatype in all datasets in an integer.

@@

HDF5 file format:
An HDF5 file is a self-describing format which combines data and metadata. High level objects are stored alongwith metadata that is used to describe various characteristics about the object, alongwith the relationship between various objects.

Parallel HDF5 (PHDF5):
As described previously, HDF5 supports parallelism using MPI. Multiple processes can be used to create and store objects in a file. PHDF5 exports a parallel I/O interface, where internally makes use of MPI-IO to perform I/O in parallel. 
Applications typically store multiple objects in a single HDF5 file. However, many popularly used parallel file systems like Lustre, Panasas etc. are known to perform poorly for workloads where multiple processes access a shared file (commonly referred to as the N-->1 access pattern).

\subsection{PLFS}
PLFS is a middleware virtual file system that converts writes to a shared logical file into writes to multiple physical files. 
It is situated between the application and the parallel file system responsible for the actual data storage. 
It transforms an application's N-->1 access pattern into an N-->N access pattern, where every process participating in I/O writes data to its own , separate file. 
Thus, it interposes on application I/O and converts its I/O pattern into one that is more suitable for the underlying parallel file system.
The basic operation of PLFs is as follows. For every logical file, PLFS creates a /texit{container} structure on the underlying parallel file system. 
This container is a hierarchical directory tree consisting of multiple sub-directories. Multiple processes opening the same logical file for writing get a unique data file within the container. When the processes write to their file, a record identifying the write is appended to an index file shared between all processes. Thus, PLFS maintains sufficient metadata to recreate the shared logical file. A FUSE daemon exports this container structure as a single, logical file to the end-users, thereby keeping the PLFS operations transparent from them.

>>>>

\subsection{HDF5}
HDF5 is a versatile data model containing complex data objects and metadata. 
Its information set is a collection of datasets, groups, datatypes and metadata objects. The data model defines mechanisms for creating associations between various information items. The main components of HDF5 are described below.

File:
In the HDF5 data model the container of an HDF5 infoset
is represented by a file. It is a collection of objects and also explains the relationship between them.
Every file begins with a root group "/", which serves as the "starting-point" in the object hierarchy.

Dataset:
HDF5 datasets are objects that represent actual data or content. Datasets are arrays which can have multiple dimensions. A dataset is characterized by a dataspace and a datatype. The dataspace captures the rank (number of dimensions), and the current and maximum extent in each dimension. The datatype describes the type of its data elements.

Group:
A group is an explicit association between HDF5 objects. It is synonymous with directories in a file system. A group could contain multiple other groups, datasets or datatypes \textit{within} it.

Attribute:
Attributes are used for annotating datasets, groups, and datatype objects. They are datasets themselves, and are \textit{attached} to the existing objects they annotate.

For example, as shown in figure~\ref{hdf5_example}, the file "Sample.h5" contains the root group which itself contains a group G1 and two datasets, D1 and D2. Group G1 contains a dataset D3. Attribute A1 is linked to dataset D1. The objects and the relationships between them can be represented as a B-tree, which is used internally by HDF5 to index its objects.

@@

%Parallel HDF5 (PHDF5):
%HDF5 supports parallelism using MPI. Multiple processes can be used to access objects in a file. PHDF5 exports a parallel I/O interface, internally making use of MPI-IO. 
%Applications typically store multiple objects in a single HDF5 file. However, many popularly used parallel file systems like Lustre, Panasas etc. are known to perform poorly for workloads where multiple processes access a shared file.

Virtual Object Layer (VOL):
The Virtual Object Layer (VOL) is a new abstraction layer internal to the HDF5 library~\cite{vol}. It is implemented just below the public API. The VOL exports an interface that allows writing plugins for HDF5, thereby enabling developers to store objects in a format different from the default HDF5 file format (like native netCDF or HDF4 format). Plugin writers provide an implementation for a set of functions that access data on disk. These include functions for file management, dataset creation and access, group creation, to name a few.

\subsection{PLFS}
PLFS is a middleware virtual file system that converts writes to a shared logical file into writes to multiple physical files. 
It is situated between the application and the parallel file system responsible for the actual data storage. 
It transforms N-1 into N-N, where every process participating in I/O writes data to its own, separate file. 
The basic operation of PLFS is as follows. For every writer to a logical file, PLFS creates a unique physical file on the underlying parallel file system. 
It also maintains sufficient metadata to recreate the shared logical file. 
We added a new feature to PLFS called extendible attributes (Xattrs). Xattrs serve as short, extensible metadata stored as key-value pairs.
They can be used to store user-defined information about data for easy and fast retrieval. 
%We use xattrs to store metadata about HDF5 datasets. For example, the datatype, number of dimensions and their extent are stored as xattrs.

Users can interface with PLFS directly by using the PLFS API or by using an MPI library that has been integrated with PLFS (ad\_plfs layer).

### 2012-09-21 hdf5/discs_2012/evaluation.tex
almost complete. yet to change figures

For evaluation purposes, we have used HDF5's \textit{h5perf} performance tool~\cite{h5perf}. h5perf is a tool for testing the performance of PHDF5. It allows configuring various parameters, such as the number of processes to run the test with, number of datasets to be created in the file, amount of data contributed by each process per dataset, amount of data read/written by a process in a single I/O call (transfer size) etc. For our measurements, we created 10 datasets in the file, and the total file size was 64 GB or more.

All tests were performed on the Lustre parallel file system~\cite{lustre} on the Atlas cluster at University of Dresden. The file system has 12 OSTs with a stripe size of 1 MB. The file system is connected to the compute nodes using an SDR Infiniband link. 
The cluster has 92 nodes with 64 cores each and 64 to 512 GB memory. All nodes are AMD Opteron 6274 running at 2.2 GHz.

Our tests were configured such that we spawn a maximum of 4 processes per node. Tests run with less than 4 processes utilized a single node.

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{4pes_interleaved_1m10_coll_writes}
%\caption{A sample HDF5 file}
%\label{writes}
%\end{figure}
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{4pes_interleaved_1m10_coll_reads}
%\caption{A sample HDF5 file}
%\label{reads}
%\end{figure}

>>>>

For evaluation purposes, we have used HDF5's \textit{h5perf} performance tool~\cite{h5perf}. It allows configuring various parameters, such as the number of processes to run the test with, number of datasets to be created in the file, amount of data contributed by each process per dataset, amount of data read/written by a process in a single I/O call (transfer size) etc. For our measurements, we created 10 datasets in the file, and the total file size was 64 GB or more.

All tests were performed on the Lustre parallel file system~\cite{lustre} on the Atlas cluster at University of Dresden. The file system has 12 OSTs with a stripe size of 1MB. The file system is connected to the compute nodes using an SDR Infiniband link. It should be noted that the file system is in production and subject to demanding and varying workload. We present here the average bandwidth values of 3 runs, though in certain cases, we have seen significant variation.
The cluster has 92 AMD Opteron nodes with 64 cores each and 64 to 512 GB memory.

We have performed tests with 1,2,4,8,32 and 64 processes. We spawned a maximum of 4 processes per node.
Reads and writes are either contiguous or interleaved; processes either access contiguous locations in file or execute a strided pattern.
We compare the performance of the default MPI-IO driver, our plugin, and the PLFS MPI-IO driver (ad\_plfs).

In Figure~\ref{write_contig}, we show the write performance for a transfer size of 1MB for contiguous writes, for varying number of processes. It can be seen that the plugin regularly outperforms MPI-IO. However, as we increase the number of processes, its performance falls below that of ad\_plfs. Figure~\ref{read_contig} shows the performance of contigous reads for an unaligned transfer size. The plugin again outperforms MPI-IO, but the performance of ad\_plfs is the best for 64 processes. 
%Figure~\ref{write_interleaved} shows the performance of unaligned writes for a maximum of 8 processes. We set the transfer size here to be 10 bytes more than 1MB. Also, the writes here are interleaved as opposed to contiguous. We see a similar pattern, where the plugin outperforms MPI-IO, but is not as good as that of ad\_plfs for 8 processes. 
Figure~\ref{write_collective} compares the performance of MPI-IO with that of the plugin for collective, unaligned, interleaved writes for a maximum of 8 processes. It is important to note that the plugin outperforms MPI-IO in collective mode. We have not been able to obtain results for collective ad\_plfs due to resource restrictions.

Overall, our results show that although the plugin generally shows good performance, it does not scale as well as we increase the number of processes. This is due to the fact that since we are directly making calls to the PLFS API, the metadata overhead seems to be high. We have all processes participating in file open and close operations, which for large problem sizes can add siginificant overhead. Hence, for future work, we plan to use ad\_plfs directly in the plugin which can overcome this performance drawback, since ad\_plfs has collective optimizations in open and close. It should however be noted that MPI libraries like Volpex-MPI~\cite{volpex}, which do not have MPI-IO as part of their MPI library, could make use of a plugin that does not rely on MPI-IO.

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4PEs_node_1M_contig_ind_writes}
\caption{Performance of contiguous writes}
\label{write_contig}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4pes_node_1m_ind_read}
\caption{Performance of contiguous reads}
\label{read_contig}
\end{figure}

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{interleaved_1M10_4pes_writes_noncoll}
%\caption{Performance of interleaved writes}
%\label{write_interleaved}
%\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4pes_interleaved_1m10_coll_reads}
\caption{Performance of collective writes}
\label{write_collective}
\end{figure}

### 2012-09-21 hdf5/discs_2012/introduction.tex
almost complete. yet to change figures

Hierarchical Data Format (HDF5) is a data model , library and file format for storing and managing data. It is designed for flexible and efficient I/O. HDF5 defines an information set, which is a container of array variables, groups, and types. The data model defines mechanisms for creating associations between various information items. HDF5 is widely used in the industry and scientic domain, in understanding global climate change, special effects in film production, DNA analysis, weather prediction, financial data management etc. (citation here)
Parallel HDF5 (PHDF5) enables developing high performance, parallel applications using standard technologies like MPI in conjunction with HDF5. An HDF5 file created using PHDF5 is compatible with serial HDF5 files and is shareable between different platforms. PHDF5 exports a standard parallel I/O interface which then uses MPI's parallel I/O functionality.

The central component of HDF5, the file, is a self-describing format which combines data and metadata. Users typically store multiple HDF5 objects in a single file, and the library stores metadata alongwith data that describes the relationships between various objects amongst other metadata. 
However, this native single-file format has its disadvantages. A PHDF5 application has multiple processes accessing a single HDF5 file. Many popular parallel file systems are known to behave poorly under these circumstances. Secondly, since many HDF5 objects are stored in a single file, this eliminates any possibility of performing useful semantic analysis on objects beyond the scope of an HDF5 application. In this paper, we address these shortcomings of HDF5, viz. performance issues due to multiple processes accessing a single shared file, and lack of a way to perform useful post-processing on HDF5 objects inherent due to the native file format.

PLFS is a middleware virtual file system developed at Los Alamos National Lab (LANL). It converts writes to a shared logical file into writes to multiple physical files.  
Thus, it interposes on application I/O and converts its I/O pattern into one that is more suitable for the underlying parallel file system.
PLFS is popularly used as a checkpoint file system where applications demonstrating heavy checkpoint workloads are known to benefit significantly from PLFS.

In this work, we have developed a new plugin for HDF5 using its recently introduced Virtual Object Layer (VOL). The VOL exports an interface that allows writing plugins for HDF5, where it primarily passes references to HDF5 objects, thereby enabling developers to store objects in a format different from the default HDF5 file format. 
We have written a plugin with two main objectives. The plugin stores data in a unique way that enables semantic post-processing on HDF5 objects outside the scope of the HDF5 application. Secondly, PLFS converts N-->1 accesses into N-->N accesses, thereby showing a significant improvement in performance as well. Preliminary results using HDF5's h5perf performance tool show a Xx performance improvement in I/O overall traditional PHDF5.

The rest of the paper is organized as follows. We represent details about HDF5 and PLFS in section 2, describe our plugin design and implementation in section 3, and our evaluation in section 4. We present related work in section 5, the current status of the plugin alongwith future work in section 6, and finally we conclude in section 7.

>>>>

Hierarchical Data Format (HDF5) is a technology suite for efficient management of large and complex data~\cite{hdf5_ad11}. It supports complex relationships between data and dependencies between objects. HDF5 is widely used in industry and scientic domains, in understanding global climate change, special effects in film production, DNA analysis, weather prediction, financial data management etc.~\cite{hdf5_users}
Parallel HDF5 (PHDF5)~\cite{phdf5} enables developing high performance, parallel applications using standard technologies like MPI in conjunction with HDF5. 
%An HDF5 file created using PHDF5 is compatible with serial HDF5 files and is shareable between different platforms. 
PHDF5 exports a standard parallel I/O interface which itself uses MPI's parallel I/O functionality. This is used alongwith parallel file systems to achieve high performance I/O.

An HDF5 file is a self-describing format which combines data and metadata. It is a container in which users typically store multiple HDF5 objects alongside their metadata. 
However, this native single-file format has its disadvantages. A PHDF5 application has multiple processes accessing a single file (i.e. N-1 access pattern). Unfortunately, many popular parallel file systems are known to behave poorly under these circumstances~\cite{lustre_perf}\cite{nersc_io}. Secondly, since many HDF5 objects are stored in a single file, this complicates the possibility of performing useful semantic operations on individual objects. In this paper, we enhance the performance of HDF5 by addressing the performance issues inherently arising from poor file system performance on N-1 access patterns, and augment performing  useful post-processing on HDF5 objects.

PLFS (Parallel Log-Structured File System) is a middleware virtual file system developed at Los Alamos National Lab (LANL)~\cite{plfs_sc09}. It converts writes to a shared logical file into writes to multiple physical files to overcome the performance bottleneck associated with N-1 writes. However, it suffers from an inability to understand the structure of the data that it stores.

In this work, we have developed a new plugin for HDF5 using its recently introduced Virtual Object Layer (VOL).  
This plugin serves two purposes: it stores data in a unique way that enables semantic post-processing on HDF5 objects, and by using PLFS to convert N-1 accesses into N-N accesses, thereby resulting in improved I/O performance.

%The rest of the paper is organized as follows. We present some background on HDF5 and PLFS in section 2, our plugin design and implementation in section 3, our evaluation in section 4, and our conclusion in section 5.

### 2012-09-21 hdf5/discs_2012/plugin.tex
almost complete. yet to change figures

The plugin stores data in a format which enables performing semantic analysis on the data. 
It stores every HDF5 object in a separate file/directory, so that data from different objects is not contained in the same file. 
The central object store in HDF5, the HDF5 file, is represented by a directory. Groups are stored as directories as well, whereas datasets, which contain raw data, are stored in files. 
The relationship between objects is explained by the relative paths between the objects at the file system level. As an example, the HDF5 object D1 which in HDF5's B-tree format can be represented as /G1/G2/D1 is stored on the file system as /path-to-hdf5-file/G1/G2/D1. This way , the plugin eliminates the need to store the metadata describing the relationship between objects. Metadata about datasets, such as the datatype, extent, dimensions etc., are stored as PLFS Xattributes (xattrs) as explained in the previous section.

Consider the example shown figure 1. Using the plugin, file test.h5 is stored as a directory. Groups G1, G2, G3 are represented as directories at the paths /tmp/test.h5/G1, /tmp/test.h5/G2 , /tmp/test.h5/G1/G3/ respectively. Datasets D1 and D2 are files stored under their respective groups.

This approach gives us the ability to distinguish HDF5 objects at the file system level, without requiring an HDF5 application to provide information about objects and the relationship between them. 
We can now perform post-processing and semantic analysis on the data, outside the scope of the HDF5 application.

>>>>

We have developed a plugin which stores data in a format which enables performing semantic analysis on the data and uses PLFS for high I/O throughput. 
Instead of storing all objects in a single file, it stores every HDF5 object in a separate location, so that data from different objects is not contained in the same file. In short, we provide a raw mapping of the HDF5 objects to the file system. 
HDF5 files and groups are stored as directories, whereas datasets and attributes, which contain raw data, are stored in files created using PLFS. 
Consider the example shown in figure~\ref{hdf5_example}. Using the plugin, file Sample.h5 is stored as a directory. Group G1 is stored as a directory under it, and datasets D1 and D2 are files. Attribute A1 is a file created at the same location as dataset D1 to which it is attached. Additionally, the name of the attribute is stored as \textit{dataset name}.\textit{attribute name} to denote the dataset to which the attribute is attached. Thus these objects are stored at the following paths:\\
/Sample.h5/\\
/Sample.h5/G1\\
/Sample.h5/G1/D3\\
/Sample.h5/D1\\
/Sample.h5/D1.A1\\
/Sample.h5/D2\\

We can see that the relationship between the objects is represented by their relative paths at the file system. That is, the path /Sample.h5/G1/D3 tells us that D3 is a dataset belonging to Group G1 under the root group of the file. This way, the plugin also eliminates the need to store the metadata describing the relationship between objects. Metadata about datasets, such as the datatype, extent, dimensions etc. are stored as Xattrs. 
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{hdf5_example_plugin}
%\caption{HDF5 file created using our PLFS plugin}
%\label{hdf5_example_plugin}
%\end{figure}

This approach gives us the ability to distinguish HDF5 objects inside the storage system.

Our plugin makes direct calls to the PLFS API. For the current version of our plugin, we do not support collective I/O operations. It should be noted that HDF5 when used natively with MPI-IO allows users to specify whether collective I/O should be used for reading and writing data.

The above format of storing data allows us to perform atleast two different analysis optimizations.  To illustrate via an example, imagine storing a three-dimensional ocean model within a PLFS file. The storage system sees the file as an opaque linear array of bytes.  With the structure, however, PLFS can provide \textit{active analysis} as well as \textit{semantic restructuring}.

Active analysis borrows the transducers idea from the Semantic File System~\cite{semantic_fs} which has since been productized in BigTable and Hbase~\cite{google_coprocessors,GFS,apache_hbase}.  With active analysis, the application can ship a data parser function when it creates the PLFS file.  As the data is written into PLFS, PLFS can apply the data function on the streaming data.  The function will output key-value pairs which PLFS can embed in its extensible metadata.  In this example, one simple function might record the height of the largest wave.  Due to PLFS's model of storing a logical file across multiple physical files, the PLFS extensible metadata can record the height of the largest wave within each physical file.  However, given that PLFS now understands the structure of the logical file, these multiple physical files, within the PLFS container, are more accurately thought of as shards.  In a future burst buffer architecture~\cite{burst_buffers}, these semantic shards will be spread across multiple burst buffer nodes.  Therefore, subsequent analysis of the ocean model can quickly find the burst buffer containing the shard with the largest wave by searching a small amount of extensible metadata instead of scanning the entire ocean model.

Semantic restructuring is the idea of reorganizing the data into a new set of semantic shards.  This would be done to speed future analysis routines.  For example, assume that the ocean model was originally sharded using a row-order organization (i.e. across latitude instead of longitude).  An analysis routine which will explore the model along a column-ordering will suffer poor performance with the row-order organization as its access pattern will result in a large number of small reads from a large set of semantic shards.  However, by knowing the semantic structure, the analysis routine can request a semantic restructuring which will be a compact, intuitively described request such as "restructure into row-ordering."  Without structural knowledge, a semantic restructuring would be significantly more complicated: the analysis routine would have to send a large list of logical offsets to PLFS to inform it of expected read patterns.  In an exascale system, the list of logical offsets will be in the order of one billion.  Semantic restructuring shrinks the size of the request to a small constant value.
